Bees with larger brains thrive in urban environmnets

### 2023-02-05 Manuscript/Word.Rmd
Preparing the MS for biology letters, formatting on length, number of figures and adding pertinent refs

---
title: "Bees need larger brains to thrive in urban environmnets"
output:
  pdf_document:
    fig_caption: yes

link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt

bibliography: 
  - references.bib

header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float} 
- \floatplacement{figure}{H}

csl: BiologyLetters.csl

---

@@

\textbf{Jose B. Lanuza$^{1}$} \orcidlink{0000-0002-0287-409X} \textbf{, Miguel Á. Collado$^{2}$}  \orcidlink{0000-0002-4216-317X} \textbf{, Ferran Sayol$^{3}$} \orcidlink{0000-0003-3540-7487} \textbf{and Ignasi Bartomeus$^{1}$} \orcidlink{0000-0001-7893-4389}

$^{1}$Estación Biológica de Doñana (EBD-CSIC), E-41092 Seville, Spain

$^{2}$Departamento de Ciencias de la Computación e Inteligencia Artificial, Universidad de Sevilla, Seville, Spain

$^{3}$Centre de Recerca Ecològica i Aplicacions Forestals: Bellaterra, Catalonia, Spain

\textbf{Author for correspondence: barragansljose@gmail.com}

@@

# Abstract

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role in ecosystems for fruit and seed production. However, not all pollinators are negatively affected by habitat conversion, but some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. While this hypothesis has received ample support in studies of vertebrates, it is less obvious whether the small brain of insects can offer enough behavioural plasticity to cope with the many challenges of anthropogenic habitats. To address this issue, we measured brains for `r sum(d$individuals)` individuals from `r nlevels(factor(unique(d$Species)))` species of bees, the most important group of pollinators, and compared them with detailed information on habitat preferences. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size, but those who prefer urban habitats tended to have larger relative brain size. In addition, urban bees also tend to be larger, and hence, show larger absolute brain sizes. These results suggest that a large brain can help maintaining bee urban populations.

Keywords: relative brain size, habitat preference, Apoidea, urbanization.

[REFEREBCES TO ADD: https://link.springer.com/article/10.1007/s10980-017-0592-x; https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2435.12486; https://link.springer.com/article/10.1007/s10531-020-02003-8; https://pubmed.ncbi.nlm.nih.gov/28976620/ ; https://www.biorxiv.org/content/10.1101/2022.12.09.519739v1 (this one for big body size in cities in disc. also see https://onlinelibrary.wiley.com/doi/10.1111/eva.13087); https://onlinelibrary.wiley.com/doi/10.1111/ele.14133?af=R (maybe in the intro); https://www.frontiersin.org/articles/10.3389/fevo.2020.00058/full]()

@@

Human-dominated habitats ---notably cities--- drastically modify the original conditions were pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to enjoy the benefits of an urban life. In a novel environment, the chances to successfully survive and reproduce largely depends on cognition, that is, the processes involved in gathering, storing and reacting to environmental information. Cognitive limitations have thus help explaining the low success of many species in novel environments in general, and urban areas in particular. For example, reptiles, amphibians and birds that thrived after being introduced to new regions outside their native ranges tend to have larger brains (relative to their body size) than unsuccessful invaders [@sol2008; @sol2005; @amiel2011]. [maybe also a sentence on urban dwelers having larger brains? Or if there are good references, replace the concept of invaders, by the one on urban species]() However, similar evidence is lacking for insects, in part because advanced cognitive abilities were not considered possible with their tiny brains until recently [citar a chitka]()[Que refs?](J). We now know that bees and other insects have advanced cognition, being able for instance to use tools to resolve new problems and learn socially how to do so [@alem2016]. Which role plays bees cognitive abilities to adapt to anthropocentric habitats has so far not being explored.

@@

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. For North America, we selected states located on the East Coast of United States (Figure S1A) that covered an approximate area of `r usakm2` km$^{2}$.  Similarly, we selected countries located on the north and center of Europe (Figure S1B) that represented a total area of `r eukm2` km$^{2}$. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a distribution larger than the area selected; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications, which is from 2006.

@@

## (d) Construction of the phylogenetic tree

We constructed a phylogeny with all studied species by using a previously published genus-level bee phylogeny [@hedtke2013]. For this, we first read the backbone phylogeny with the function *read.tree* and drop the missing genera with function *drop.tip* from the package *ape* version 5.6-2 [@paradis2019]. Then, we selected as an outgroup a wasp genera (*Tachysphex*) and recalculated distances to the root with the function *chronos* from *ape*. Species tips were added to genera nodes as polytomies of equal branch length relative to the genera branch length using the phytools package version 1.2-0 [@revell2012]. From the resulting tree, we created a scaled phylogenetic covariance matrix with the function *inverseA* from the package *MCMCglmm* version 2.33 [@hadfield2010]. Finally, we estimated the phylogenetic signal of relative brain size and habitat preference across habitats (see next section) for our set of species with the help of the function *phylosig* from the *phytools* package.

## (e) Analysis

@@

All our analyses were undertaken in R version 4.0.5 [@r2021], and all data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

Bees with a preference for inhabiting urban habitats tended to have bigger brains related to their body sizes, and those with forest or agricultural  preferences had relatively smaller brains. This pattern is highly suggestive, because investing in brain tissue is expensive (REF), and hence it only payoff in unpredictable habitats where a flexible behavior is needed. When the environment is stable, small relative brains are cost-effective to exploit predictable resources, but in highly unpredictable environments, a large brain can confer the needed flexibility to exploit unpredictable resources. Indeed, specialist bees also tend to have larger brains [@sayol2020], as they need to locate and memorize patchy resources. This same complex learning and navigation skills may be also needed in urban environments. Urban habitats are typically harsher, with less and/or new resources to explore, also containing new unknown threats and a highly changing environment [@wheater2002]. Hence, urban dwellers may need to have better cognitive abilities and flexibility in their behavior than forest or agricultural bees.

Some examples of behaviours needed to survive in urban environments are the use of human-made materials to nest [@allasino2019] or being able to collect pollen and nectar from exotic flowers from gardens [@lowenstein2019]. We should not expect forested habitats to require such a plastic behaviour because the resources found in these habitats are homogeneous, stable, and bees are already pre-adapted to use them [@smith2019]. In addition, other traits have already been described to be filtered out in cities, such as smaller body sizes (REF), early spring species or brood parasites [@harrison2019]. We also found small species to be less frequent in urban habitats, maybe as a consequence of the patchy resource distribution, which need from larger body sizes, which confer larger foraging ranges, to be accessed [@kendall2019].

Despite brain weight is relatively easy to measure and standardize with the used protocol, recent studies challenge the use of brain size as an estimator for behavioral complexity and computational power [@chittka2009\; and references therein]. However, many studies have shown correlations between brain size and performance [@deaner2007; @gronenberg2010; @sol2008], and brain size is correlated with learning abilities in bees [@collado2021]. Finer scale measures such as neuropils size or detailed mushroom bodies measurements [the suspected centers of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] could further explain differences in behaviours, and in consequence, shed extra light on species habitat associations. In any case, raw measures of brain size have the advantage to allow for larger comparisons across species.

@@

# Data accessibility

All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.5s5kuf. In addition, bee trait data and code can be found in the online repositories of Zenodo XXXXXXXKXX and Github https://github.com/ibartomeus/declines_brain.git.

# Authors' contributions

JBL, MAC, FS, IB...

# Funding

# Acknowledgements

# Conflict of interest declaration

We declare we have no competing interests.

@@

# Supplementary material

## Brain and body size measurements

Brain size was measured as the weight of the fixed brain for each specimen. First, all collected individuals were kept in cold and sent to the laboratory where they were identified by expert taxonomists and anesthetized in cold. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.

We considered as a proxy of body size the average intertegular span (ITS) per species. ITS represents the distance between the base of the wings (tegulae) on the bee thorax and has been shown to accurately predict body size as it is highly correlated with dry body weight (Cane et al., 2006; Kendall et al., 2019). All ITS measurements were conducted with a stereomicroscope (magnification 16x or 80x) with a calibrated ocular micrometre (resolution down to 0.02 mm).

Cane, J. H., Minckley, R. L., Kervin, L. J., Roulston, T. A. H., & Williams, N. M. (2006). Complex responses within a desert bee guild (Hymenoptera: Apiformes) to urban habitat fragmentation. _Ecological applications_, _16_(2), 632-644.

Kendall, L. K., Rader, R., Gagic, V., Cariveau, D. P., Albrecht, M., Baldock, K. C., ... & Bartomeus, I. (2019). Pollinator size and its consequences: Robust estimates of body size in pollinating insects. _Ecology and Evolution_, _9_(4), 1702-1714.

```{r Table S1, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center"}

#Create table of occurrences per habitat type (USA)
#Load libraries
library(tidyverse)
library(kableExtra)

#Load extracted data
pref <- read_csv("../Data/Usa_data/land_cover_usa.csv.gz") %>% 
dplyr::select(species, cover.names) %>% 
mutate_if(is.character,as.factor)

#Create table with original and new cover categories
pref %>% 
rename(Cover_names = cover.names) %>% 
select(Cover_names) %>% 
group_by(Cover_names) %>% 
summarise(Occurrences = length(Cover_names)) %>% 
mutate(Cover_names1 = fct_recode(as.factor(Cover_names),
       Discarded = "Barren Land", 
       Agricultural = "Cultivated Crops", 
       Natural = "Deciduous Forest", 
       Urban = "Developed, High Intensity", 
       Urban = "Developed, Medium Intensity",
       Urban = "Developed, Low Intensity",
       Urban = "Developed, Open Space", 
       Natural = "Emergent Herbaceous Wetlands", 
       Natural = "Evergreen Forest", 
       Agricultural = "Hay/Pasture", 
       Natural = "Herbaceous", 
       Natural = "Mixed Forest", 
       Discarded = "Open Water", 
       Natural = "Shrub/Scrub", 
       Natural = "Woody Wetlands")) %>% 
arrange(desc(Occurrences)) %>%  
kbl(col.names = c("Original NLCD Cover class", "Occurrences", "Assigned cover class"), caption = "\\textbf{Table S1.} Original downloaded cover classes for United States with the respective assigned cover classes and number of bee occurrences found on each of them.",
booktabs = T, align = c("l", "c", "l"), linesep = "") %>% 
kable_styling(font_size = 12, latex_options = c("hold_position", "striped")) %>% 
row_spec(0,bold=TRUE)

```{r Table S2, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center"}

#Create table of occurrences per habitat type (USA)
#Load libraries
library(tidyverse)
library(kableExtra)

#Load extracted data
pref <- read_csv("../Data/Europe_data/land_cover_europe.csv.gz") %>% 
    dplyr::select(Species, Cover_names) %>% 
    mutate_if(is.character,as.factor)

#Create table with original and new cover categories
pref %>% 
select(Cover_names) %>% 
group_by(Cover_names) %>% 
summarise(Occurrences = length(Cover_names)) %>% 
mutate(Cover_names1 = fct_recode(as.factor(Cover_names),
      Urban = "Airports", #1
      Discarded = "Bare rocks", #2
      Natural = "Beaches, dunes, sands", #3
      Natural = "Broad-leaved forest", #4
      Natural = "Coastal lagoons", #5
      Agricultural = "Complex cultivation patterns", #6
      Natural = "Coniferous forest", #7
      Urban = "Construction sites", #8
      Urban = "Continuous urban fabric", #9
      Urban = "Discontinuous urban fabric", #10
      Discarded = "Dump sites", #11
      Discarded = "Estuaries", #12
      Agricultural = "Fruit trees and berry plantations", #13
      Urban = "Green urban areas", #14)
      Urban = "Industrial or commercial units", #15
      Natural = "Inland marshes", #16
      Natural = "Intertidal flats", #17
      Agricultural = "Land principally occupied by agriculture, with significant areas of natural vegetation", #18
      Discarded = "Mineral extraction sites", #19
      Natural = "Mixed forest", #20
      Natural = "Moors and heathland", #21
      Agricultural = "Natural grasslands", #22
      Agricultural = "Non-irrigated arable land", #23
      Agricultural = "Pastures", #24
      Natural = "Peat bogs", #25
      Urban = "Port areas", #26) 
      Discarded = "Road and rail networks and associated land", #27
      Natural = "Salt marshes", #28
      Discarded = "Sea and ocean", #29
      Natural = "Sparsely vegetated areas", #30
      Urban = "Sport and leisure facilities", #31
      Natural = "Transitional woodland-shrub", #32
      Agricultural = "Vineyards", #33
      Discarded = "Water bodies", #34
      Discarded = "Water courses" #35  
)) %>% 
mutate(Cover_names = fct_recode(as.factor(Cover_names),
      "Land principally occupied by agriculture" = "Land principally occupied by agriculture, with significant areas of natural vegetation")) %>% 
arrange(desc(Occurrences)) %>%  
kbl(col.names = c("Original CLC Cover class", "Occurrences", "Assigned cover class"), caption = "\\textbf{Table S2.} Original downloaded cover classes for Europe with the respective assigned cover classes and number of bee occurrences found on each of them.",booktabs = T,
align = c("l", "c", "l"), linesep = "") %>% 
kable_styling(font_size = 12,latex_options = c("hold_position", "striped")) %>% 
row_spec(0,bold=TRUE)

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.align="center", fig.height=12, fig.width=12, fig.cap="\\textbf{Figure S1.} Maps showing the study areas and the number of GBIF occurrences in logarithmic scale for the selected bee species. A) East Coast of United States with records for Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. B) European continent with records downloaded for England, Belgium, Netherlands, Germany and Denmark.",  out.width="90%"}

#Plot maps with density of bee occurrences

#Europe----
#Load libraries
library(giscoR) #map of europe
library(ggplot2)
library(dplyr)
library(readr)
library(sf)
library(patchwork)
library(rnaturalearth)

#Load europe map
Europe <- gisco_get_countries(region = "Europe")
#Load countries filtered
euro_map <- st_read("../Data/Europe_data/euro_map_1.shp", quiet = TRUE)
#Load data with records filtered
all_long_lat <- read.table("../Data/Europe_data/all_above_50_europe.csv.gz",  header=T, quote="\"", sep=",")

#Plot Europe map with counties of interest highlighted
p1 = ggplot(Europe) +
geom_sf(fill="gray25",color = "black", alpha = 0.5,size=0.2)  +
geom_sf(data= euro_map %>% group_by(CNTR_CODE) %>% summarise() ,
aes(fill = CNTR_CODE, group=CNTR_CODE),fill = "#f2ea69", color = "black",alpha=0.8, size = 0.25) +
coord_sf(xlim = c(-10, 30), ylim = c(35, 65))  + 
microplot::theme_collapse(panel.border = element_rect(colour = "black", fill=NA, size=2),
panel.background = element_rect(fill = "white"))

p2 = ggplot(Europe) +
geom_sf(fill="gray80",color = "black", alpha = 1,size=0.2)+
geom_bin2d(data = all_long_lat, aes(x = long, y = lat, fill=log(..count..)),size = 1,bins=70) +
viridis::scale_fill_viridis(name = "log(Count)", breaks = c(0, 4, 8), limits=c(0,8)) +
coord_sf(xlim = c(-8, 18), ylim = c(40, 60)) +
theme(panel.background = element_rect(fill = "white"),
panel.border = element_rect(colour = "black", fill=NA, size=1)) +
ylab("Latitude") +
xlab("Longitude") +
ggtitle("B") +
theme(legend.position = c(0, 1),legend.justification = c(0, 1),
legend.background = element_blank(),
legend.box.background = element_blank(),
legend.key = element_blank(), 
legend.title = element_text(face = "bold"),
axis.title = element_text(face="bold", size = 14),
plot.title = element_text(face="bold", size = 18),
legend.text = element_text(face = "bold", size = 12), legend.direction="horizontal") +
scale_y_continuous(breaks = seq(40, 60, by = 6)) +
scale_x_continuous(breaks = seq(-8, 18, by = 8)) +
guides(fill = guide_colourbar(title.position = "top", title.hjust = .5)) +
ggspatial::annotation_scale(location = 'br',  bar_cols = c("grey60", "white"),  width_hint = 0.15)

#Insert general plot within specific plot
panel1 = p2 + inset_element(p1, left = 0, bottom = -0.89 , right = 0.35, top = 1.23, ignore_tag = TRUE)

#USA-----
#Load libraries

#Load data
usa_states_1 <- read.csv("../Data/Usa_data/usa_states.csv")
ny_metropolitan_1 <- read.csv("../Data/Usa_data/ny_metropolitan.csv")
all_long_lat <- read.table("../Data/Usa_data/all_above_50_usa.csv.gz",  header=T, quote="\"", sep=",")

#Load maps
canada <- ne_states(country = "canada", returnclass = "sf") %>%
st_as_sf(crs = 4326)

usa <- ne_states(country = "united states of america", returnclass = "sf") %>%
st_as_sf(crs = 4326)

p1 = ggplot(canada) + 
geom_sf(fill="gray60",color = "black", alpha = 1,size=0.2) + 
geom_sf(data= usa, fill="gray60",color = "black", alpha = 1,size=0.2) + 
microplot::theme_collapse(panel.border = element_rect(colour = "black", fill=NA, size=2),
panel.background = element_rect(fill = "white")) +
geom_map(data = ny_metropolitan_1, 
map = ny_metropolitan_1,aes(long, lat, map_id = region, group = group), fill = "#f2ea69",color = "black",
size = 0.1) +
geom_map(data = usa_states_1, map = usa_states_1,
aes(long, lat, map_id = region, group = group), color = "black", fill = "#f2ea69",
size = 0.1) +
coord_sf(xlim = c(-86, -62), ylim = c(25, 48))

p2 = ggplot(canada) + 
geom_sf(fill="gray80",color = "black", alpha = 1,size=0.2) + 
geom_sf(data= usa, fill="gray80",color = "black", alpha = 1,size=0.2) +
geom_bin2d(data = all_long_lat, aes(x = long, y = lat, fill=log(..count..)),size = 1,bins=50) +
viridis::scale_fill_viridis(name = "log(Count)", breaks = c(0, 4, 8), limits=c(0,8)) +
coord_sf(xlim = c(-82, -67), ylim = c(33, 47))  +
theme(panel.background = element_rect(fill = "white"),
panel.border = element_rect(colour = "black", fill=NA, size=1)) +
ylab("Latitude") +
xlab("Longitude") +
ggtitle("A") +
theme(legend.position = c(0, 1),legend.justification = c(0, 1),
legend.background = element_blank(),
legend.box.background = element_blank(),
legend.key = element_blank(),
legend.title = element_text(face = "bold"),
axis.title = element_text(face="bold", size = 14),
plot.title = element_text(face="bold", size = 18),
legend.text = element_text(face = "bold", size = 12, hjust=0), legend.direction="horizontal") +
scale_y_continuous(breaks = seq(33, 47, by = 4)) +
scale_x_continuous(breaks = seq(-82, -67, by = 4)) +
guides(fill = guide_colourbar(title.position = "top", title.hjust = .5))+ 
ggspatial::annotation_scale(location = 'br',  bar_cols = c("grey60", "white"))

#Insert general plot within specific plot
panel2 = p2 + inset_element(p1, left = 0, bottom = -0.86, right = 0.35, top = 1.2, ignore_tag = TRUE)

library(patchwork)
panel2  + panel1

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=7, fig.cap="\\textbf{Figure S2.} Association between the log-transformed values of brain and body size (N = 116). The different raw values or points are coloured by the subfamily taxonomic rank.", fig.pos="p", out.width="80%"}

#Plot encephalization
#Load libraries
library(stringr)
library(readr)
library(ggplot2)
library(dplyr)
library(elementalist) #devtools::install_github("teunbrand/elementalist")

#Read data
m = read_csv("../Data/Processing/residuals_model.csv")
r = read_csv("../Data/Processing/residuals_data.csv")
d = read_csv("../Data/Processing/taxonomic_rank.csv") %>% 
mutate(Species = str_replace_all(query, " ", "_")) %>% 
select(!c(query, db))

#Add taxonomic rank to dataset
r = left_join(r,d)

#Add ggplot
ggplot(m, aes(x = log_it, y = estimate__)) +
geom_line(size=1) +
geom_point(data = r, aes(log_it, log_brain_weight, color = subfamily), size=3.5) + 
theme_bw() +
ylab("log(Brain size)") +
xlab("log(Body size)") +
theme(legend.position = c(.95, .60),
legend.justification = c("right", "top"),
legend.box.just = "right",
legend.background = element_rect_round(radius = unit(0.2, "snpc"),fill = "grey85"),
legend.key = element_rect(fill = "grey85")) +
guides(colour=guide_legend(title="Subfamily")) +
theme(axis.title = element_text(face = "bold"))

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=5, fig.width=12, fig.cap="\\textbf{Figure S3.} Frequency distribution of the bee preferences for each habitat type (i.e., natural, agricultural and urban). The different habitats are reclassifications from the 2006 land use categories of the Corine Land Cover (CLC) inventory for Europe and the National Land Cover (NLC) database for Unites States.", fig.pos="p"}

#Histogram of habitat preferences
#Load libraries
library(tidyverse)

#Load preferences
preferences_europe = read_csv("../Data/Europe_data/preferences_europe.csv") 
preferences_usa = read_csv("../Data/Usa_data/preferences_usa.csv")

#Merge datasets
preferences = bind_rows(preferences_europe, preferences_usa)

#Convert to long to model everything at the same time
long_data = preferences %>% gather(Habitat, Preference, 2:4, -c(Species))

#Relevel for plotting
long_data = long_data %>% 
mutate(Habitat=fct_relevel(Habitat,c("Natural","Agricultural", "Urban")))

labels.minor <- c("Avoidance", rep("", times = 2), "Preference","")

long_data$Habitat <- factor(long_data$Habitat, levels = c("Natural", "Agricultural", "Urban"),labels = c("A) Natural", "B) Agricultural", "C) Urban"))

long_data = long_data %>% 
mutate(Preference_colour = case_when(Preference <= 0.2 ~ "A", Preference >= 0.8 ~ "B", T ~ "C"))

long_data$Preference_colour <- factor(long_data$Preference_colour, levels = c("A","C", "B"), labels = c("Low preference", "No preference", "High preference"))

ggplot(long_data, aes(x=Preference, color= Preference_colour)) +
    geom_histogram(aes(fill= Preference_colour),bins = 35, alpha = 0.8,linewidth=0.5) + 
    xlab("Habitat preference") +
    ylab("Frequency") +
    facet_wrap(~ Habitat) +
    scale_fill_manual(name="" ,values=c("coral2", "gray", "palegreen3")) +
    scale_color_manual(name="" ,values=c("coral2", "gray", "palegreen3")) +
    theme_bw() +   
    theme(axis.title = element_text(face = "bold", size = 14),
    panel.border=element_rect(size=1),
    strip.text = element_text(face = "bold", size = 18,hjust = -0.01),
    strip.background = element_rect(fill = NA, colour = NA, size = 0,linetype="solid"),
    panel.grid.minor=element_blank())+
    scale_y_continuous(expand = c(0, 0), limits = c(0,60))+
    geom_vline(aes(xintercept = 0.20), colour="black", linetype="dashed") +
    geom_vline(aes(xintercept = 0.80), colour="black", linetype="dashed") +
    scale_x_continuous(breaks = seq(from = 0, to = 1, by = 0.2),
                       labels = c(0,0.2,0.4,0.6,0.8,1)) +
  theme(legend.position = 'bottom', legend.box.background = element_rect(colour = "black",size=1))

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=8, fig.cap="\\textbf{Figure S4}. Association between relative brain size (A and D), brain weight (B and E) and intertegular distance (C and F) with habitat preference separated by habitat type (i.e., natural, agricultural and urban) and geographical region (USA in the upper panel and Europe in the lower one). The shaded and coloured areas by habitat type represent 95\\% credible intervals.", fig.pos="p", out.width="100%"}

plot_all = readRDS("../Data/Processing/plot_all.rds")

plot_all

>>>>

---
title: "Bees with larger brains thrive in urban environmnets"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt
bibliography: references.bib
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}
---

@@

@@

@@

Human-dominated habitats ---notably cities--- drastically modify the original conditions were pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to enjoy the benefits of an urban life. In a novel environment, the chances to successfully survive and reproduce largely depends on cognition, that is, the processes involved in gathering, storing and reacting to environmental information. Cognitive limitations have thus help explaining the low success of many species in novel environments in general, and urban areas in particular. For example, reptiles, amphibians and birds that thrived after being introduced to new regions outside their native ranges tend to have larger brains (relative to their body size) than unsuccessful invaders [@sol2008; @sol2005; @amiel2011].  However, similar evidence is lacking for insects, in part because advanced cognitive abilities were not considered possible with their tiny brains until recently. We now know that bees and other insects have advanced cognition, being able for instance to use tools to resolve new problems and learn socially how to do so [@alem2016]. Which role plays bees cognitive abilities to adapt to anthropocentric habitats has so far not being explored.

@@

## Overview

To evaluate our main hypothesis, we have used a dataset of bee species captured in North America and Europe with measurements of brain mass and body size [@sayol2020]. For all these species, we downloaded the occurrence records from the Global Biodiversity Information Facility (GBIF). After filtering the data to avoid possible biases, we extracted the land use classification of the different georeferenced records and summarised them into 3 different categories: (1) natural, (2) agricultural and (3) urban. To evaluate if the different bee species occur more in some habitats than others, we calculated their habitat preference by comparing the observed occurrences with the ones of null models assuming random distributions across habitats [@collado2019]. Finally, for the species that showed low and high statistical preference for the different habitats, we investigated how the association between habitat preference and relative brain size changed by habitat type.

All analyses were undertaken in R version 4.0.5 [@r2021] and the manuscript was written in rmarkdown version 2.17 [@markdown2022]. All data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. We selected the boundaries from the maps packages *spData* version 2.2.0 [@spdata] and *giscoR* vesion 0.3.2 [@giscoR], and extracted the occurrences with the help of the function *st_intersection* from the package *sf* version 1.0-7 [@pebesma2018]. For North America, we selected states located on the East Coast of United States (Figure S1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure S1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.5s5kuf where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a distribution larger than the area selected; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications, which is from 2006. After pruning the data, our final dataset comprised `r spp_number` species with  `r usa_occurrences` occurrence records for United States (`r usa_spp` species; median occurrences = `r usa_median`) and `r eu_occurrences` records for Europe (`r eu_spp` species; median occurrences = `r eu_median`) wit a total of `r spp_shared` shared species between both regions.

@@

## (d) Analysis

@@

@@

Bees with a preference for inhabiting urban habitats tended to have bigger brains related to their body sizes, and those with forest or agricultural  preferences had relatively smaller brains. This pattern is highly suggestive, because investing in brain tissue is expensive (REF), and hence it only payoff in unpredictable habitats where a flexible behavior is needed. When the environment is stable, small relative brains are cost-effective to exploit predictable resources, but in highly unpredictable environments, a large brain can confer the needed flexibility to exploit unpredictable resources. Indeed, specialist bees also tend to have larger brains [@sayol2020], as they need to locate and memorize patchy resources. This same complex learning and navigation skills may be also needed in urban environments. Urban habitats are typically harsher, with less and/or new resources to explore, also containing new unknown threats and a highly changing environment (Wheater, 1999). Hence, urban dwellers may need to have better cognitive abilities and flexibility in their behavior than forest or agricultural bees.

Some examples of behaviours needed to survive in urban environments are the use of human-made materials to nest (Allasino et al., 2019) or being able to collect pollen and nectar from exotic flowers from gardens (Lowenstein et al., 2014). We should not expect forested habitats to require such a plastic behaviour because the resources found in these habitats are homogeneous, stable, and bees are already pre-adapted to use them (Smith et al., 2019). In addition, other traits have already been described to be filtered out in cities, such as smaller body sizes (REF), early spring species or brood parasites (Harrison et al., 2017). We also found small species to be less frequent in urban habitats, maybe as a consequence of the patchy resource distribution, which need from larger body sizes, which confer larger foraging ranges, to be accessed [@kendall2019].

Despite brain weight is relatively easy to measure and standardize with the used protocol, recent studies challenge the use of brain size as an estimator for behavioral complexity and computational power (Chittka and Niven, 2009, and references therein). However, many studies have shown correlations between brain size and performance (Deaner et al., 2007; Gronenberg and Couvillon, 2010; Lefebvre and Sol, 2008), and brain size is correlated with learning abilities in bees (Collado et al 2021). Finer scale measures such as neuropils size or detailed mushroom bodies measurements (the suspected centers of cognitive processes in bees; Barth & Heisenberg, 1997; Withers et al., 1995; Fahrbachand & Robinson 1995) could further explain differences in behaviours, and in consequence, shed extra light on species habitat associations. In any case, raw measures of brain size have the advantage to allow for larger comparisons across species.

@@

@@
